---
title: "Getting started with y2conjoint"
vignette: >
  %\VignetteIndexEntry{Getting started with y2conjoint}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
library(y2conjoint)
```

`y2conjoint` provides data structures and functions for working with the output
of a choice-based conjoint study. This article walks through a full example:
loading raw utilities, describing their structure, defining products, and
estimating preference shares for a competitive set.

## The raw data

A conjoint model produces one *part-worth utility* per respondent per attribute
level. These arrive in a terse "model format" where columns are named
`A[NUM]B[NUM]` — attribute *i*, level *j*. The package ships a small simulated
example:

```{r}
#| label: raw-data
head(example_utilities)
```

Alongside the utilities we need a **crosswalk** that says what each column means:
which attribute (collection) it belongs to, its human-readable name, and — for
ordered attributes like price — its rank.

```{r}
#| label: crosswalk
example_crosswalk
```

Here `Brand` is unordered (`collection_order` is `NA`), while `Price` and
`Battery` are ordered.

## Building a conjoint_df

`conjoint_df()` renames the model-format columns to their user-facing names and
records the collection structure as metadata. Extra columns (IDs, demographics)
are carried along untouched.

```{r}
#| label: build
cjt <- conjoint_df(example_utilities, example_crosswalk)
cjt
```

The print header summarizes the collections (a trailing `*` marks an ordered
one), the outside-good column, and how many extra columns are present.

### Collections

Each collection groups the level columns of one attribute:

```{r}
#| label: collections
cjt |>
  protected_cols()
```

The level columns and the `NONE` outside good are **protected**. You can wrangle
demographics freely, but the utility columns cannot be dropped, renamed, or
overwritten by accident.

```{r}
#| label: filter-ok
# Filtering rows and editing demographics is fine.
cjt |>
  dplyr::filter(age >= 40) |>
  dplyr::mutate(age_decade = age %/% 10 * 10)
```

```{r}
#| label: protected-error
#| error: true
# Overwriting a protected column is blocked.
cjt |>
  dplyr::mutate(Northwind = 0)
```

If you genuinely need to reshape the utilities, call `tibble::as_tibble()` first
to drop the guardrails.

## Describing products with specs

A `spec` describes one product by selecting levels across collections. Selecting
a single level per collection is the common case:

```{r}
#| label: spec
flagship <- spec(cjt, c("Meridian", "$399", "20 hours"), name = "Flagship")
flagship
```

You may select *several* levels within one collection — useful for co-branded
products, where the combine function (below) decides how to merge them.

```{r}
#| label: spec-cobrand
co_brand <- spec(cjt, c("Northwind", "Cascade", "$199", "10 hours"), name = "Co-brand")
```

## Competitive sets and preference shares

A `competitive_set` bundles the products that compete for choice. `run_scenario()`
applies a logit (softmax) choice model: it turns each respondent's product
utilities into choice probabilities, lets the `NONE` outside good compete, and
averages the probabilities into mean preference shares.

```{r}
#| label: scenario
market <- competitive_set(
  spec(cjt, c("Northwind", "$199", "20 hours"), name = "Value"),
  spec(cjt, c("Cascade", "$299", "20 hours"), name = "Mainstream"),
  spec(cjt, c("Meridian", "$399", "20 hours"), name = "Premium"),
  name = "Launch"
)

shares <- run_scenario(cjt, market)
shares
```

`total_share` is the combined share captured by the three products; the
remainder goes to the outside good.

```{r}
#| label: plot
#| fig-alt: "Bar chart of preference share by product."
#| fig-width: 6
#| fig-height: 3
library(ggplot2)

shares |>
  dplyr::select(-total_share) |>
  tidyr::pivot_longer(everything(), names_to = "product", values_to = "share") |>
  dplyr::mutate(product = sub("^share_", "", product)) |>
  ggplot(aes(share, reorder(product, share))) +
  geom_col() +
  labs(x = "Preference share", y = NULL)
```

## Combining levels within a collection

When a product selects more than one level from a collection, `combine_fn`
decides how to merge them. The default, `pmax()`, takes the best available level
(a reasonable rule for co-branding: buyers respond to the more appealing brand).

You can also pass a **named list** to use different rules per collection. Any
collection you omit falls back to `pmax()`:

```{r}
#| label: combine-fn
run_scenario(
  cjt,
  market,
  combine_fn = list(Brand = pmax, Price = pmax)
)
```

## Scaling choice probabilities

`scaling_factor` multiplies every utility before the softmax. Values above 1
sharpen the shares toward the most attractive product; values below 1 flatten
them toward an even split. It is a convenient handle for calibrating a simulator
to known market conditions.

```{r}
#| label: scaling
run_scenario(cjt, market, scaling_factor = 2)
```
